---
title: "US_elections"
author: "Max Derie, Ignacio Pascale, Ishita Gupta, Oana Muntean, Xinyu Yu"
date: "27-11-2020"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(vroom)
library(tidyquant)
library(ggplot2)
library(ggrepel)
library(scales) 
library(tidytext)
library(ggridges)
library(sf)
library(hrbrthemes)
library(urbnmapr)
library(showtext)
library(ggrepel)
library(patchwork)
library(plotly)


font_add_google("Courier Prime", "Courier")
showtext_auto()

```

# Data Cleaning

We will first import the dataset `counties_fixed.csv`


```{r, message= FALSE, warning=FALSE}
election <- vroom::vroom("data/counties_fixed.csv") %>% 
  clean_names() %>% 
  mutate(winner = case_when(
    percentage20_donald_trump >= percentage20_joe_biden ~ "Trump",
    percentage20_donald_trump <  percentage20_joe_biden ~ "Biden"),
    
    winner16 = case_when(
    percentage16_donald_trump >= percentage16_hillary_clinton ~ "Trump",
    percentage20_donald_trump <  percentage16_hillary_clinton ~ "Clinton")
    )
glimpse(election)
```

Now we want to import the dataset `electoral_college.csv`. Here we have information about how many electoral votes each state has. We will filter by 2020.

```{r, warning=FALSE, message=FALSE}
# Data set with Electoral votes from all the years

electoral_votes<- read_csv(here("data","electoral_college.csv")) %>% clean_names() %>%
  filter(year==2020)
```


Let's look for duplicates

```{r, warning=FALSE, message=FALSE}
#remove empty rows and columns
temp<-remove_empty(election, which = c("rows","cols"))

#there were no empty rows to be removed
rm(temp)

#check for duplicates
#we look for entries on the same day and for the same country
election%>%get_dupes(county, state)
electoral_votes %>% get_dupes(state,year)
```

Get POLYGONS from `urbanmaps` for counties in the US

```{r}
counties_sf <- get_urbn_map("counties", sf = TRUE)
```


```{r}
counties_sf <- counties_sf %>% 
  mutate(county_name2 = case_when(
    grepl('County$', county_name) ~ str_sub(county_name, end = -8),
    grepl('Parish$', county_name) ~ str_sub(county_name, end = -8),
    TRUE ~ county_name)
  )
```

Join with `elections` dataset.

```{r}
data <- counties_sf %>% 
  left_join(election, by = c( "county_name2" = "county",  "state_abbv" = "state"))
glimpse(data)
```

View datasets summary.

```{r}
# Summary
summary(election)
summary(data)
```


Now we will create a new dataset with the STATE POLYGONS from `urbanmap`  and the data from `electoral_votes`. We will add new data such as: 

- `winner_by`: who won the state in 2020
- `winner_by16`: who won the state in 2016
- `was_flipped`: whether the state was flipped in 2020


We will use this data to produce maps and analyze the underlying trends on them. The data included was taken from [this article](https://www.theguardian.com/us-news/ng-interactive/2020/dec/02/us-election-results-2020-joe-biden-defeats-donald-trump-to-win-presidency)


```{r}
# Identified swing states
swing_states <- c("Arizona","Texas","Florida","Georgia","Pennsylvania","Ohio","Wisconsin","North Carolina","South Carolina","Iowa","Nevada","Michigan")
# Swing states Won by Trump
trump_won_swing<- c("Texas","Florida","Ohio","North Carolina","South Carolina","Iowa")
# Swing states won by Biden
biden_won_swing<- c("Arizona","Georgia","Pennsylvania","Wisconsin","Nevada","Michigan")
# Swing states won by Trump in 2016
trump_won_swing16<-c("Arizona","Texas","Florida","Georgia","Pennsylvania","Ohio","Wisconsin","North Carolina","South Carolina","Iowa","Michigan")
# Swing states won by Hillary in 2016
hillary_won_swing16<-c("Nevada")
# Swing states flipped by Biden in 2020
flipped<-c("Arizona","Georgia","Pennsylvania","Wisconsin","Michigan")



electoral_mapping<- 
  get_urbn_map("states", sf=TRUE) %>%
  left_join(electoral_votes %>% select(!year), 
            by =c("state_name" = "state")
            ) %>%
  
  # Create variable winner_by
  mutate(winner_by = ifelse(state_name %in% trump_won_swing, 
                            "Trump", 
                            ifelse(state_name %in% biden_won_swing, 
                                   "Biden", NA)),
         # Create variable winner_by 16
         winner_by16 = ifelse(state_name %in% trump_won_swing16, "Trump", 
                              ifelse(state_name %in% hillary_won_swing16, "Hillary", NA)),
         # Create variable was flipped
         was_flipped = ifelse(state_name %in% flipped, TRUE, FALSE),
         # Create variable colour
         colour = ifelse(state_name %in% trump_won_swing, "red",
                         ifelse(state_name %in% hillary_won_swing16, "blue",
                                ifelse(state_name %in% flipped, "flipped",NA))))
```


# Visualizations

## General Overview

Battle Trump against Biden


```{r}
party_colours <- c("Biden" = "#2E74C0", "Trump"= "#CB454A")

election %>%
  select(votes20_donald_trump,votes20_joe_biden) %>% 
  drop_na() %>% 
  summarise(Biden = sum(votes20_joe_biden),
            Trump = sum(votes20_donald_trump)) %>% 

  ggplot() + 
  geom_col(aes(x = 1, y = 538, fill = "Trump"), width = 1) +
  geom_col(aes(x = 1, y = 306, fill = "Biden"), width = 1) +
  scale_fill_manual(values = party_colours) + 
  
  labs(title = "Biden triumphs by 6M votes",
       subtitle = "2020 election vote counts",
       fill = NULL) + 
  theme_void() +
  theme(legend.position = "top",
         plot.title = element_text(family = "Courier", face = "bold", size = 20),
         plot.subtitle = element_text(family = "Courier", size = 12),
         legend.text=element_text(family="Courier"),
         legend.title=element_text(family="Courier"))+
  xlim(c(0,2)) + 
  coord_flip()  +
  geom_linerangeh(aes(y = 270, xmin = 0.5, xmax = 1.75),
                 size = 0.7,
                 color = "black") +
  annotate("text", x=1.82, y=270, label="270 electors", size = 4,color = "black",family = "Courier") +
  # annotate("text", x=1.6, y=70, label="306 electors", size = 8,color = "#2E74C0") +
  # annotate("text", x=1.6, y=470, label="232 electors", size = 8,color = "#CB454A") +
  annotate("text", x=1, y=35, label="306", size = 10,color = "white", fontface = "bold", family = "Courier") +
  annotate("text", x=1, y=505, label="232", size = 10,color = "white", fontface = "bold",family = "Courier") +
  annotate("text", x=0.4, y=70, label="78M votes (52%)", size = 5,color = "#2E74C0",family = "Courier") +
  annotate("text", x=0.4, y=465, label="72M votes (48%)", size = 5,color = "#CB454A",family = "Courier") +
  
  theme(plot.title = element_text(face = "bold"))
  
```

Representation of counties


```{r}
party_colours <- c("Biden" = "#2E74C0", "Trump"= "#CB454A")

p1 <- election %>%
  select(votes20_donald_trump,votes20_joe_biden) %>% 
  drop_na() %>% 
  summarise(Biden = sum(votes20_joe_biden),
            Trump = sum(votes20_donald_trump)) %>% 

  ggplot() + 
  geom_col(aes(x = 0.76, y = Biden + Trump, fill = "Trump"), width = 1.5) +
  geom_col(aes(x = 0.76, y = Biden, fill = "Biden"), width = 1.5) +
  scale_fill_manual(values = party_colours) + 
  
  labs(title = "Counties do not display people",
       subtitle = "2020 election vote counts",
       fill = NULL) + 
  theme_void() +
  theme(plot.title = element_text(family = "Courier", face = "bold", size = 20),
         plot.subtitle = element_text(family = "Courier", size = 12),
        legend.position = "top",
        text = element_text(family = "Courier"))+
  xlim(c(0,2)) + 
  coord_flip()  +
  # geom_linerangeh(aes(y = 0.5*(Biden+Trump), xmin = 0.5, xmax = 1.75),
  #                size = 0.5,
  #                color = "black") +
  
  annotate("text", x=0.76, y=12000000, label="78M", size = 10,color = "white", fontface = "bold",family = "Courier") +
  annotate("text", x=0.76, y=137500000, label="72M", size = 10,color = "white", fontface = "bold",family = "Courier") +
  annotate("text", x=1.8, y=18000000, label="By vote count:", size = 4.5,color = "black", fontface = "bold",family = "Courier") +
  
  theme(plot.title = element_text(face = "bold"))
  
```



```{r, warning=FALSE, message=FALSE}
party_colours <- c("Biden" = "#2E74C0", "Trump"= "#CB454A")

summarised_el <- election %>%
  select(votes20_donald_trump,votes20_joe_biden) %>% 
  drop_na() %>% 
  summarise(Biden = sum(votes20_joe_biden),
            Trump = sum(votes20_donald_trump))
p2 <- data %>%
  # select(votes20_donald_trump,votes20_joe_biden) %>% 
  # drop_na() %>% 
  # pivot_longer(cols = Biden:Trump, names_to = ("candidate")) %>% 
  ggplot() + 
  geom_bar(aes(x = 0.76, fill = winner), position = position_stack(reverse = T), width = 1.5) + 
  # geom_col(`dataaes(x = 1, y = Biden + Trump, fill = "Trump"), width = 1)
  # geom_bar(aes(x = 1, y = Biden + Trump, fill = "Trump")) +
  # geom_col(aes(x = 1, y = Biden, fill = "Biden")) +
  scale_fill_manual(values = party_colours, na.translate = F) + 
  # labs(title = "Trump gets majory of counties dispite loss",
  #      subtitle = "2020 election vote counts by counties",
  #      fill = NULL) + 
  theme_void() +
  theme(plot.title = element_text(face = 'bold'),
        legend.position = "top",
        text = element_text(family = "Courier"))+
  xlim(c(0,2)) + ylim(c(0,3115)) +
  coord_flip() +
  annotate("text", x=0.76, y=210, label="515", size = 10,color = "white", fontface = "bold",family = "Courier") +
  annotate("text", x=0.76, y=2800, label="2600", size = 10,color = "white", fontface = "bold",family = "Courier") +
  annotate("text", x=1.8, y=320, label="By counties:", size = 4.5,color = "black", fontface = "bold",family = "Courier") +
  guides(fill = FALSE)

library(patchwork)
p1/p2
  
``` 

Votes in 2016


```{r}
# electoral votes battle
party_colours <- c("Clinton" = "#2E74C0", "Trump"= "#CB454A")

election %>%
  select(votes20_donald_trump,votes20_joe_biden) %>% 
  drop_na() %>% 
  summarise(Biden = sum(votes20_joe_biden),
            Trump = sum(votes20_donald_trump)) %>% 

  ggplot() + 
  geom_col(aes(x = 1, y = 538, fill = "Trump"), width = 1) +
  geom_col(aes(x = 1, y = 232, fill = "Clinton"), width = 1) +
  scale_fill_manual(values = party_colours) + 
  
  labs(title = "Trump won by electoral vote difference",
       subtitle = "2016 Election Results",
       fill = NULL) + 
  theme_void() +
  theme(legend.position = "top",
         plot.title = element_text(family = "Courier", face = "bold", size = 20),
         plot.subtitle = element_text(family = "Courier", size = 12),
         legend.text=element_text(family="Courier"),
         legend.title=element_text(family="Courier"))+
  xlim(c(0,2)) + 
  coord_flip()  +
  geom_linerangeh(aes(y = 270, xmin = 0.5, xmax = 1.75),
                 size = 0.7,
                 color = "black") +
  annotate("text", x=1.82, y=270, label="270 electors", size = 4,color = "black",family = "Courier") +
  # annotate("text", x=1.6, y=70, label="306 electors", size = 8,color = "#2E74C0") +
  # annotate("text", x=1.6, y=470, label="232 electors", size = 8,color = "#CB454A") +
  annotate("text", x=1, y=35, label="232", size = 10,color = "white", fontface = "bold", family = "Courier") +
  annotate("text", x=1, y=505, label="306", size = 10,color = "white", fontface = "bold",family = "Courier") +
  annotate("text", x=0.4, y=70, label="66M votes (48%)", size = 5,color = "#2E74C0",family = "Courier") +
  annotate("text", x=0.4, y=465, label="63M votes (46%)", size = 5,color = "#CB454A",family = "Courier") +
  
  theme(plot.title = element_text(face = "bold"))

```


Votes by Counties


```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=8}
data %>% 
  ggplot(aes()) +
  geom_sf(aes(fill = winner, colour = winner)) +
  scale_fill_manual(values = c("#2E74C0","#CB454A"), na.translate = F) +
  scale_colour_manual(values = c("#2E74C0","#CB454A"), na.translate = F) +
  coord_sf(datum = NA) +
  theme_void() +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "top",
        text = element_text(family = "Courier")) + 
  labs(title = "US turns red despite Trump loss",
       subtitle = "US county map colored by winner",
       fill = NULL) +
  guides(color= FALSE)
```

Now we will create circles to visualize this by population

```{r}
circles_data <- data
st_geometry(circles_data) <- NULL

circles_data_sf <- circles_data %>%
  drop_na(long, lat) %>%
  filter(state_abbv != "HI") %>% 
  mutate(long = long,
         lat = lat) %>% 
  st_as_sf(coords = c('long', 'lat'),
           crs = 4326)

glimpse(circles_data_sf)

library("openxlsx")
# Write the first data set in a new workbook
write.xlsx(circles_data, 'electcion.xlsx')
```

Land doesn't vote, people do.

```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=8}
ggplot() +
  geom_sf(data = data %>% filter(state_abbv != "AK", state_abbv != "HI"), fill = "#F5F5F5", colour = "grey70") +
  geom_sf(data = circles_data_sf, aes(colour = winner, size = total_pop),key_glyph = "rect") + 
  scale_colour_manual(values = c("#2E74C0","#CB454A"), na.translate = F) +
  scale_size_area(max_size = 7) + #scale proportional to population
  coord_sf(datum = NA) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "top",
        text = element_text(family = "Courier")) + 
  labs(title = "Land doesn't vote - people do",
       subtitle = "County winners scaled to population size") +
  guides(color=guide_legend(NULL), size = FALSE)
```
## Swing States Data

Identify Swing States

```{r identify_swing_states, warning=FALSE, message=FALSE, fig.width=8}

colour_scale<- c("grey","#8A0303")
fill_scale<- c("white","#8A0303")


identify_swing<- data %>%
  # Create abbreviations in data
  mutate(swing = ifelse(state_name %in% swing_states, TRUE, FALSE),
         name_swing = ifelse(swing == TRUE, state_abbv, NA)) %>%
  
  # Group by
  group_by(state_name, swing, name_swing) %>%
  summarise(long = max(long),
            lat = max(lat)) %>%
  
  # Plot
  ggplot() +
  # Geom SF
  geom_sf(aes(fill = swing, colour=swing)) +
  # Fill Scale
  scale_fill_manual(values =  fill_scale) +
  # Colour Scale
  scale_colour_manual(values = colour_scale)+
  
  # Geom SF Text
  geom_sf_text(aes(label = name_swing), 
               colour="white", 
               family = "Courier",
               face= "bold",
               size=5) +
 
   # Remove Coordinates
  coord_sf(datum = NA) + 
  # Theme Void
  theme_void() +
  
  # Add Labels
  labs(
    title = "Which are the American Swing States?", 
    subtitle = "Where the election actually happened") +
  
  # Theme Settings
  theme(
    legend.position = "none",
    text = element_text(family= "Courier",size=15),
    plot.title = element_text(face="bold")
    
    ) +
  
  NULL


identify_swing
```

How many electoral votes do Swing States have?

```{r electoral_votes_swing, warning=FALSE, message=FALSE, fig.height=6, fig.width=6}

plot_votes<- electoral_votes %>%
  filter(state %in% swing_states)%>%
  arrange(votes) %>% mutate(
    cumulative = cumsum(votes)
  ) %>%
  
  ggplot() +
  geom_col(aes(x = reorder(state, cumulative), y = votes), fill="steelblue") +
  geom_line(aes(x = reorder(state, cumulative), group=1, y = cumulative)) +
  geom_point(aes(x =reorder(state, cumulative), y=cumulative)) + 
  
  # Add arrow
  
   geom_segment(aes(x = 12, y = 194, xend = 8, yend = 193),
                  arrow = arrow(length = unit(0.5, "cm")), colour="#8A0303") +
  
  # Annotate
  
  annotate("text", x = 7.5, y = 193, label = '194', size = 5, angle = 0, fontface = "bold") +
  
  # Labels
  
  labs(title = "Swing States add up to 194 electoral votes",
       subtitle = "Cumulative increase of electoral votes in Swing States",
       x = "Swing States",
       y = "Electoral Votes") + 
  
  # Theme settings
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 20),
        text = element_text(family="Courier", size=15),
        plot.title = element_text(face="bold"))

plot_votes
    
  
```


Swing States - Outcome in 2016

```{r, warning=FALSE, message=FALSE, fig.width=8}
electoral_mapping %>%
  mutate(colour = replace_na(colour, "none")) %>%
  
  
  ggplot() +
  geom_sf(aes(fill=winner_by16, colour=winner_by16))+
  scale_fill_manual(values = c("#2E74C0","#CB454A" ), na.translate=FALSE) +
  scale_colour_manual(values = c("#2E74C0","#CB454A"), na.value= "grey80") +
  guides(colour = FALSE) +
  geom_sf_text(aes(label=ifelse(!is.na(winner_by16),votes," ")), color='white', family="Courier", size=5) +
  coord_sf(datum=NA) +
  theme_void() +
  labs(title = "Trump won 188 electoral votes from Swing States in 2016",
       subtitle = "Outcome in Swing States in 2016 Elections",
       fill = "Winner") +
  
  theme(text = element_text(family="Courier", size=15),
        plot.title = element_text(face="bold")) + 
  NULL

```

Swing States Flipped by Biden

```{r, warning=FALSE, message=FALSE, fig.width=8}
electoral_mapping %>% 
  
  
  ggplot() +
  geom_sf(aes(fill=was_flipped, colour=was_flipped))+
  scale_fill_manual(values = c("white", "#2E74C0")) +
  scale_colour_manual(values = c("grey", "#2E74C0")) +
  geom_sf_text(aes(label=ifelse(was_flipped==TRUE,state_abbv," ")), colour="white",size=5, family = "Courier") +
  #geom_sf_label(aes(label=ifelse(was_flipped==TRUE,electoral_votes," ")),size=3.75, family = "Courier") +
  coord_sf(datum=NA) +
  theme_void() +
  labs(title = "Joe Biden Flipped 5 States in the Elections",
       subtitle = "Flipped States from 2016 to 2020") +
  
   theme(legend.position = "none",
         text = element_text(family="Courier", size=15),
        plot.title = element_text( face="bold")) +
  
  
  NULL
```

Swing States Flipped by Trump

```{r trump_flipped, warning=FALSE, message=FALSE, fig.width=8}

electoral_mapping %>%
  mutate(trump_flipped = "States Flipped by Trump") %>%
  
  ggplot() +
  geom_sf(aes(fill=trump_flipped, colour=trump_flipped))+
  scale_fill_manual(values = c("white")) +
  scale_colour_manual(values = c("grey")) +
  guides(fill = FALSE) +
  geom_col(aes(x=0, y=0), fill="#CB454A", show.legend = TRUE) +
  coord_sf(datum=NA) +
  theme_void() +
  
  labs(title = "Donald Trump was not persuasive enough",
       subtitle = "Mr President did not manage to flip any states from 2016 to 2020",
       colour = " ")+
  theme(text = element_text(family = "Courier", size= 15),
        plot.title = element_text(face="bold")) +
  NULL

```

Map - Swing States Got Biden The Presidency

```{r general_view, warning=FALSE, message=FALSE, fig.width=8}
electoral_mapping %>%
  mutate(#was_flipped = replace_na(was_flipped,FALSE),
        # winner_by = replace_na(winner_by, "none"),
         colour = replace_na(colour, "none")) %>%
  
  
  ggplot() +
  geom_sf(aes(fill=winner_by, colour=colour, group=was_flipped), size=.75)+
  scale_fill_manual(values = c("#2E74C0","#CB454A" ), na.translate = FALSE) +
  scale_colour_manual(values = c("#2E74C0",muted("#fffb0a"),"grey80","#CB454A")) +
  geom_sf_text(aes(label=ifelse(was_flipped==TRUE,votes," ")), color='white', family="Courier", size=5) +
  coord_sf(datum=NA) +
  theme_void() +
  #theme(legend.position = "none") + 
  guides(colour = FALSE)+
  labs(title = "The Flipped Swing States Got Biden The Presidency",
       subtitle="Flipped Swing States Highlighted Made The Difference in 2020",
    fill = "Winner"
  ) +
  
    theme(text = element_text(family="Courier",size=15),
        plot.title = element_text(face="bold")) +
  
  NULL





```

Bar Plot - Swing States Got Biden The Presidency

```{r, fig.width= 9, fig.height=9}
# final electoral vote battle

flipped<-c("Arizona","Wisconsin","Michigan","Pennsylvania","Georgia")

plot_votes_2<- electoral_votes %>%
  filter(state %in% flipped)%>%
  arrange(votes) %>% mutate(
    cumulative = cumsum(votes)
  ) %>%
  
  ggplot() +
  geom_col(aes(x = reorder(state, cumulative), y = votes), fill="steelblue") +
  geom_line(aes(x = reorder(state, cumulative), group=1, y = cumulative)) +
  geom_point(aes(x =reorder(state, cumulative), y=cumulative)) + 
  
  # Add arrow
  
  geom_segment(aes(x = 5, y = 20, xend = 5, yend = 73),
                  arrow = arrow(length = unit(0.5, "cm")), colour="#8A0303") +
  
  # Arrow heading down
  
   geom_segment(aes(x = 5, y = 73, xend = 5, yend = 20),
                  arrow = arrow(length = unit(0.5, "cm")), colour="#8A0303") +
  
  # Annotate
  
  annotate("text", x = 4.5, y = 50, label = '73 Votes', size = 13, angle = 0, fontface = "bold") +
  
  # Labels
 # geom_text_repel(aes(x = states, y = votes, label=votes)) +
    labs(title = "The Swing States Got Biden The Presidency",
       subtitle = "Cumulative increase of electoral votes in Swing States flipped",
       x = "Swing States Flipped",
       y = "Electoral Votes") + 
  
  # Theme settings
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 20),
        text = element_text(family="Courier", size=20),
        plot.title = element_text(face="bold"))





plot_votes_2


```


# Demographics data

We will now analyze the demographics of the data.

Biden wins in Counties with more people.

```{r, population_trend}


population_data<- data %>%
  filter(total_pop<2500000) 
  
plot<- ggplot(population_data,aes(x=total_pop,y=percentage20_joe_biden))+
  geom_point(alpha=0.5,colour='blue')+
  xlab('Total population of county') +
  ylab('Percentage voted for Joe Biden')+
  labs(title="Biden wins in counties with more people!",
       subtitle ="Percentage votes to Biden by population in county") +
  theme_minimal()+
  theme(text = element_text(family="Courier"))
  #theme(plot.title = element_text(face = "bold"))
#ggtitle("Biden wins in counties with more people!")


ggplotly(plot) %>%
  layout(title = list(text = paste0('<b>','Biden wins in counties with more people!','<b>',
                                    '<br>',
                                    '<sup>',
                                    'Percentage votes to Biden by population in county',
                                    '</sup>')))

```

Did the rich vote for Trump?

```{r, income_trend}

plot_1<- ggplot(data,aes(x=income_per_cap,y=percentage20_joe_biden,color=unemployment))+
  geom_point(alpha=0.8)+
  scale_colour_gradientn(colours = terrain.colors(30))+
  xlab('Income per capita for county') +
  ylab('Percentage voted for Joe Biden')+
  theme_minimal()+
  theme(text = element_text(family="Courier"))+
  ggtitle("Did the rich vote for Trump? Not really...")

ggplotly(plot_1) %>%
  layout(title = list(text = paste0('<b>','Did the rich vote for Trump? Not really...','<b>',
                                    '<br>',
                                    '<sup>',
                                    'Percentage votes to Biden by income per capita and unemployment ',
                                    '</sup>')))
```


White people vote more for Trump?

```{r, race_trend, warning=FALSE, message=FALSE}
g<-pivot_longer(data,c(black,white),names_to="race",values_to = "percentage")

#plot_2<- ggplot(g,aes(x=percentage,y=percentage20_donald_trump,colour=race))+
plot_2<- ggplot(g,aes(x=percentage,y=percentage20_donald_trump,colour=race))+
  theme_minimal()+
  geom_point(alpha=0.5)+
  xlab('percentage of race in the county') +
  ylab('Percentage voted for Donald Trump')+
  ggtitle("White people vote more for Trump?")+
  theme(text = element_text(family="Courier"))+
  scale_color_manual(breaks = c("black", "white"),
                        values=c("blue", "red"))

ggplotly(plot_2) %>%
  layout(title = list(text = paste0('<b>','White people vote more for Trump?','<b>',
                                    '<br>',
                                    '<sup>',
                                    'Percentage votes to Trump split by race',
                                    '</sup>'))) 
#plot_2
```


But who did actually vote for trump?

```{r, profession_trend, warning=FALSE, message=FALSE}

#z<-pivot_longer(data,c(self_employed,family_work),names_to="profession_type",values_to = "percentage_prof") 

plot_3<- 
  ggplot(data,aes(x=self_employed,y=percentage20_donald_trump))+
  theme_minimal()+
  geom_point(alpha=0.5,colour='red')+
  #facet_wrap(~ profession_type) +
  xlab('percentage of self employed people in the county') +
  ylab('Percentage voted for Donald Trump')+
  ggtitle("But who did actually vote for Trump?")+
    theme(text = element_text(family="Courier"))

ggplotly(plot_3) %>%
    layout(title = list(text = paste0('<b>','But who did actually vote for Trump?','<b>',
                                    '<br>',
                                    '<sup>',
                                    'Percentage votes to Trump by self employed percentage in county',
                                    '</sup>')))
```


## Voter Turnout Rate

### Time series plot

We will import 2 new datasets.

```{r, load_data}

turnout_2016 <- read.csv("data/turnout_2016.csv") %>% 
  clean_names() 

turnout_2020 <- read.csv("data/turnout_2020.csv") %>% 
  clean_names()

#join two dataset
data_turnout <- turnout_2016 %>% 
  left_join(turnout_2020, by = c("state" = "state")) %>% 
  rename(turnout_2016 = turnout_rate.x, turnout_2020 = turnout_rate.y) %>% 
  select(state, turnout_2016, turnout_2020)

glimpse(data_turnout)

```

Highest Turnout Rate Ever

```{r, load_time_series_data, warning=FALSE, message=FALSE}
library(extrafont)
extrafont::loadfonts(device="pdf")
turnout_1920_2020 <- read.csv("data/turnout_1920_2020.csv") %>% 
  clean_names() 

highlight <- turnout_1920_2020[turnout_1920_2020$year == 2020, ]
turnout_1920_2020 %>% 
ggplot(aes(x=year, y=united_states_presidential_vep_turnout_rate))+
  geom_line(size = 1, color = "#2c7fb8") +
  geom_point(color = "#2c7fb8") +
  geom_label(data = highlight, 
                   aes(label = "66.7%"), 
                   color = "#2c7fb8",
                   box.padding = 0.25,
                   point.padding = 0.5,
             vjust = -0.5,
             hjust = 0.6,
             family = "Courier")+
  theme_minimal() +
  labs(title = "The U.S. experienced highest turnout rate in over a century",
       subtitle = "1920-2020 Voting eligible population turnout rates",
       x = "Election Year",
       y = "Turnout Rate %") +
#  geom_vline(xintercept = 2020, color = "red", linetype = 5) +
#  gghighlight::gghighlight(year == 2020, label_key = united_states_presidential_vep_turnout_rate) +
  theme_minimal() +
  scale_x_continuous(n.break = 10) +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        text = element_text(family="Courier")) +
  expand_limits(y = c(40, 80))

```

In 2020 US election, More Americans voted than in any other in over 100 years. 66.7% percent of the voting-eligible population cast a ballot, delivering the popular vote and electoral college to Joe Biden, the Democratic candidate. Despite that the pandemic introduced a number of complications to voting day, early voting and mail-in ballots brought a record turnout for some states. 

## 2016 vs 2020 turnout by state

```{r}
swing_states <- c("Arizona","Texas","Florida","Georgia","Pennsylvania","Ohio","Wisconsin","North Carolina","South Carolina","Iowa","Nevada","Michigan")
swing_win <- c( "#2E74C0", "#CB454A", "#CB454A", "#2E74C0", "#2E74C0", "#CB454A", "#2E74C0","#CB454A", "#CB454A", "#CB454A",  "#2E74C0", "#2E74C0")
swing_colors <- data.frame(swing_states, swing_win)
swing_colors1 <- swing_colors$swing_win
names(swing_colors1) <- swing_colors$swing_states
data_turnout1 <- data_turnout %>% 
  filter(state %in% swing_states) %>% 
  pivot_longer(!state, names_to = "turnout_rate", values_to = "value")
  glimpse(data_turnout1)
data_turnout1$value <- round(data_turnout1$value,1)
ggplot(data=data_turnout1,aes(x = turnout_rate, y = value, group = state)) +
  geom_line(size = 1)+
  geom_point() +
  geom_text(aes(x = turnout_rate, y = value, label = value), vjust = 2, size = 3, family = "Courier")+
  facet_wrap(~state) +
  scale_y_continuous(limits = c(40, 80)) +
  aes(color = state) +
  scale_color_manual(values = swing_colors1) +
    theme_bw()+
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold", size = 15),
          legend.position = "none",
          axis.ticks.y = element_blank(),
          axis.text.y = element_blank(),
          strip.text.x = element_text(size = 9, face = "bold"),
          text = element_text(family="Courier"))+
  scale_x_discrete(labels=c("turnout_2016" = "2016", "turnout_2020" = "2020")) +
    labs(title = "Turnout on the rise in key battlegrounds",
         subtitle = "From 2016 to 2020 presidential election")
```

Most of states saw an increase in voter turnout since 2016, with key battleground states like Florida, Michigan, Wisconsin and Pennsylvania seeing participation well above the national rate.

## Turnout map

```{r, map_turnout2020}
states_sf <- get_urbn_map("states", sf = TRUE)
data_state <- states_sf %>% 
  left_join(data_turnout, by = c("state_name" = "state"))
glimpse(data_state)
```



```{r}
data_state %>% 
  ggplot(aes()) +
  geom_sf(aes(fill = turnout_2020), colour = "#ffffff")+
  theme_void() +
  labs(title = "Turnout rate in different states",
       subtitle = "2020 Election",
       fill = "turnout rate") +
  geom_sf_text(aes(label = state_abbv), color = "white", size = 2, family = "Courier") +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 14),
        text = element_text(family="Courier")) +
  scale_fill_gradient(low = "#1a9641",
                        high = "#d7191c")
```
## Top states with highest turnout rates

```{r}
data_turnout_15 <- data_turnout %>%
  arrange(desc(turnout_2020)) %>% 
  top_n(n=15, wt = turnout_2020)
glimpse(data_turnout_15)
data_turnout_15$state
top_states <- c("Minnesota","Colorado","Maine","Wisconsin","Washington","New Hampshire","Oregon","Vermont","Michigan","Iowa","Montana","Virginia","Massachusetts","Florida","North Carolina")
top_win <- c( "#2E74C0", "#2E74C0","#2E74C0", "#2E74C0", "#2E74C0", "#2E74C0", "#2E74C0", "#2E74C0", "#2E74C0", "#CB454A", "#CB454A", "#2E74C0", "#2E74C0", "#CB454A","#CB454A")
top_colors <- data.frame(top_states, top_win)
top_colors1 <- top_colors$top_win
names(top_colors1) <- top_colors$top_states
  ggplot(data = data_turnout_15, aes(x=reorder(state, turnout_2020), y = turnout_2020)) +
  geom_point(size = 2) +
  coord_flip() +
    theme_minimal() +
    expand_limits(y = c(70, 80)) +
      labs(title = "Biden won states with high turnout rate",
           subtitle = "Top 15 states with highest turnout rates in 2020",
         y = "turnout rate")+
    theme(legend.position = "none",
        axis.title.y = element_blank(),
        plot.title = element_text(face = "bold", size = 16),
        text = element_text(family="Courier")) +
  geom_text(data=data_turnout_15, aes(x=state, y=turnout_2020, label = turnout_2020), vjust = 0.2, hjust = 1.1, color = "white") +
    aes(color = state) +
    scale_color_manual(values = top_colors1)
  
```
Among all the states, Minnesota, Colorado, Maine and Wisconsin have the highest turnout rate, and Biden won the majority of votes in those states with high voter turnout rate. 

##Correlation
Create correlation between Joe Biden and other demographic characteristics
```{r datamanip, message=FALSE, warning= FALSE, echo=TRUE}
data1<-data
data_turnout_new<-data_turnout %>% 
  filter(state!="United States") %>%
  rename(X.x=turnout_2016, X.y=turnout_2020) %>% 
  mutate(difference= (X.y-X.x)/X.x *100)
covid_data<-data1 %>% 
  group_by(state_name) %>% 
  summarise(cases= sum(cases),population=sum(total_pop), lat=round(mean(lat), 0), long =round(mean(long), 0)) %>% 
  mutate(cases_per_1M=round(cases*1000000/population,0)) 
covid_data<-st_drop_geometry(covid_data)
states <-get_urbn_map("states", sf = TRUE)
turnout_and_covid <- states %>% 
  left_join(covid_data, by = "state_name") %>% 
  left_join(data_turnout_new, by= c("state_name"= "state"))
```

```{r correlation_creation, message=FALSE, warning=FALSE}

#deactivate geomtery to mutate columns

st_geometry(circles_data_sf) <- NULL

#Create corrlation matrix
correlation <-as.data.frame(cor(circles_data_sf[, 10:54], 
                                method = "pearson", 
                                use = "complete.obs")) %>%
  #Rename empty columns
  mutate(names = rownames(.)) %>% 
  #Arrange in descending order 
  arrange(desc(percentage20_joe_biden)) %>%
  #Split into categories -Ethnicity, Economy, Job divisions and the rest (0)
  mutate(number= ifelse(names %in% c("black", "asian", "pacific", "native", "hispanic","white"),1, ifelse(names %in% c("total_pop", "employed", "income_per_cap", "poverty"),2, ifelse(names %in% c("professional", "service", "office", "family_work", "production", "self_employed", "construction"), 3, 0)))) %>% 
  filter(number !=0) %>% 
  arrange(percentage20_joe_biden) %>% 
  arrange(desc(number))%>% 
  mutate(no=row_number())
#Create 3 different datafremes for next plots
correlation1<-correlation %>% 
  filter(number ==1) 
correlation2<-correlation %>% 
  filter(number ==2) 
correlation3<-correlation %>% 
  filter(number ==3) 
  
```

```{r plot corrr, message=FALSE, warning=FALSE}
#Create first plot (for Ethnicity), reordered by correlation with Joe Biden
cor1<-ggplot(correlation1, aes(x=reorder(names, no),
                               y=percentage20_joe_biden, 
                               label=round(percentage20_joe_biden,2))) +
  #Add column names instead of numerical value
  geom_point(stat='identity', aes(col=percentage20_joe_biden), size=9)  +
  #Add gradient color in the bubble 
  scale_color_gradient2("",
                        low = muted("#CB454A"),
                        mid = "white", 
                        high = muted("#2E74C0"), 
                        midpoint=0, breaks=c(-0.5,0,0.45), 
                        labels=c("More Trump",0,"More Biden"))+ 
  
  geom_text(color="black", size=2) +
  #Show correlations between these values
  ylim(-0.6, 0.6) +
  #Change theme
  theme_minimal()+  
  #Remove x axis, change size of text and font
  theme(panel.grid.major.x= element_blank(), 
         panel.grid.minor.x = element_blank(),
         panel.background = element_blank(),
         axis.ticks=element_blank(),
        text=element_text(size=12, family="Courier"), 
        plot.title = element_text(color=muted("#CB454A"), face = "bold"))+
  
  labs(title="Blacks & Asians Voted J.Biden, People Working in Construction or Production Did Not", subtitle= "Correlation between Demographics and Votes for J.Biden",
       x="Ethnicity",
       y= "",
       fill="")+
  #Change sizes of text
  theme(plot.title = element_text(size=13), 
        plot.subtitle=element_text(size=12.5), 
        plot.caption = element_text(size=6))+
  #Flip the coordinates
  coord_flip()
#Create second plot for Economy
cor2<-ggplot(correlation2, aes(x=reorder(names, no),
                               y=percentage20_joe_biden, 
                               label=round(percentage20_joe_biden,2))) + 
  #Use stat=identity to show names
  geom_point(stat='identity', aes(col=percentage20_joe_biden), size=9)  +
  
  scale_color_gradient2("",low = muted("#CB454A"),
                        mid = "white", 
                        high = muted("#2E74C0"), 
                        midpoint=0, 
                        breaks=c(-0.5,0,0.45),
                        labels=c("More Trump",0,"More Biden"))+ 
  geom_text(color="black", size=2) +
  ylim(-0.6, 0.6) +
  #Flip coordinates
  coord_flip()+
  theme_minimal()+ 
  #Change theme 
  theme(panel.grid.major.x= element_blank(), 
         panel.grid.minor.x = element_blank(),
         panel.background = element_blank(),
         axis.ticks=element_blank(),
         text=element_text(size=12, family="Courier"))+
  #Add the category
  labs(y=NULL, x="Economy", fill="")
#Create third plot for Job division
cor3<-ggplot(correlation3, 
             aes(x=reorder(names, no), 
                 y=percentage20_joe_biden, 
                 label=round(percentage20_joe_biden,2))) + 
  
  geom_point(stat='identity', aes(col=percentage20_joe_biden), size=9)  +
  #Add diverging color from red for Trump to blue for Biden
  scale_color_gradient2("",low = muted("#CB454A"),
                        mid = "white",
                        high = muted("#2E74C0"), 
                        midpoint=0, breaks=c(-0.5,0,0.45), 
                        labels=c("More Trump",0,"More Biden"))+
  
  geom_text(color="black", size=2) +
  ylim(-0.6, 0.6) +
  coord_flip()+
  theme_minimal()+
  #Change theme
  theme(panel.grid.major.x= element_blank(), 
         panel.grid.minor.x = element_blank(),
         panel.background = element_blank(),
         axis.ticks=element_blank(),
         text=element_text(size=12, family="Courier"))+
  #Put x axis (but it will be flipped)
  labs(y="Correlation with % Votes Joe Biden", x="Job division", fill="")
#Combine plots
library(ggpubr)
#Arrange the plots on top of each other
ggarrange(cor1, cor2,cor3, ncol=1, nrow=3, align="hv", common.legend = TRUE, legend= "right", widths = c(15,1))
```
Plot correlation between covid rates and turnout rate 
```{r plotcov, echo=TRUE, message=FALSE, warning=FALSE}
turnout_and_covid%>%
  ggplot(aes(x=cases_per_1M,y= X.y)) + 
  geom_point(aes(label=state_abbv))+
  #Add line
  geom_smooth(method= "lm", se=FALSE, color=muted("#CB454A"))+
  #Change theme
  theme_minimal()+
  #Add lable titles
  labs(x="Coronavirus Cases per 1M",
       y= "Turnout Rate per State", 
       title= "More Covid less Votes!", 
       subtitle="States' Turnout Rates and Level of Covid per 1 Million People")+
  #Change font and size
  theme(text=element_text(size=10, family="Courier"), 
        plot.title = element_text(color=muted("#CB454A"), face = "bold"))
  
```
