---
title: "US_elections"
author: "Max Derie"
date: "27-11-2020"
output: html_document
---

```{r setup, include=FALSE, message FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(vroom)
library(tidyquant)
library(ggplot2)
library(ggrepel)
library(scales) 
library(tidytext)
library(ggridges)
library(sf)
library(hrbrthemes)
library(urbnmapr)
```

```{r, message= FALSE}
election <- vroom::vroom("data/counties_fixed.csv") %>% 
  clean_names() %>% 
  mutate(winner = case_when(
    percentage20_donald_trump >= percentage20_joe_biden ~ "Trump",
    percentage20_donald_trump <  percentage20_joe_biden ~ "Biden"),
    
    winner16 = case_when(
    percentage16_donald_trump >= percentage16_hillary_clinton ~ "Trump",
    percentage20_donald_trump <  percentage16_hillary_clinton ~ "Clinton")
    )

glimpse(election)
```

```{r}
#remove empty rows and columns
temp<-remove_empty(election, which = c("rows","cols"))

#there were no empty rows to be removed
rm(temp)

#check for duplicates
#we look for entries on the same day and for the same country
election%>%get_dupes(county, state)

```
```{r}
summary(election)
```


```{r}
party_colours <- c("Biden" = "#2E74C0", "Trump"= "#CB454A")

election %>%
  select(votes20_donald_trump,votes20_joe_biden) %>% 
  drop_na() %>% 
  summarise(Biden = sum(votes20_joe_biden),
            Trump = sum(votes20_donald_trump)) %>% 
  # pivot_longer(cols = Biden:Trump, names_to = ("candidate")) %>% 
  ggplot() + 
  geom_col(aes(x = 1, y = Biden + Trump, fill = "Trump")) +
  geom_col(aes(x = 1, y = Biden, fill = "Biden")) +
  scale_fill_manual(values = party_colours) + 
  
  labs(title = "Biden triumphs by 6M votes",
       subtitle = "2020 election vote counts",
       fill = NULL) + 
  theme(legend.position = "top")+
  theme_minimal() +
  xlim(c(0,2)) + 
  coord_flip() 
  
  

```



```{r}
counties_sf <- get_urbn_map("counties", sf = TRUE)
```

```{r}
counties_sf <- counties_sf %>% 
  mutate(county_name2 = case_when(
    grepl('County$', county_name) ~ str_sub(county_name, end = -8),
    grepl('Parish$', county_name) ~ str_sub(county_name, end = -8),
    TRUE ~ county_name)
  )
```

```{r}
data <- counties_sf %>% 
  left_join(election, by = c( "county_name2" = "county",  "state_abbv" = "state"))
glimpse(data)
```


```{r}
party_colours <- c("Biden" = "#2E74C0", "Trump"= "#CB454A")

data %>%
  # select(votes20_donald_trump,votes20_joe_biden) %>% 
  # drop_na() %>% 
  # pivot_longer(cols = Biden:Trump, names_to = ("candidate")) %>% 
  ggplot() + 
  geom_bar(aes(x = 1, fill = winner), position = "stack") + 
  # geom_bar(aes(x = 1, y = Biden + Trump, fill = "Trump")) +
  # geom_col(aes(x = 1, y = Biden, fill = "Biden")) +
  scale_fill_manual(values = party_colours) + 
  
  labs(title = "Trump gets majory of counties despite loss",
       subtitle = "2020 election vote counts by counties",
       fill = NULL) + 
  theme(legend.position = "top")+
  theme_minimal() +
  xlim(c(0,2)) + 
  coord_flip() 
  
```

```{r}
data %>% 
  ggplot(aes()) +
  geom_sf(aes(fill = winner), colour = "#ffffff") +
  scale_fill_manual(values = c("#2E74C0","#CB454A"))
  # scale_fill_gradientn(colours = my_scale)
```


```{r}
circles_data <- data
st_geometry(circles_data) <- NULL

circles_data_sf <- circles_data %>%
  drop_na(long, lat) %>%
  filter(state_abbv != "HI") %>% 
  mutate(long = long,
         lat = lat) %>% 
  st_as_sf(coords = c('long', 'lat'),
           crs = 4326)

glimpse(circles_data_sf)

library("openxlsx")
# Write the first data set in a new workbook
write.xlsx(circles_data, 'electcion.xlsx')
```

```{r}
ggplot() +
  geom_sf(data = data, fill = "grey70", colour = "#ffffff") +
  geom_sf(data = circles_data_sf, aes(colour = winner, size = total_pop)) + 
  scale_colour_manual(values = c("#2E74C0","#CB454A")) +
  scale_size_area() #scale proportional to population
```

```{r}
data %>% 
  ggplot(aes(child_poverty, percentage20_donald_trump)) + 
  geom_point()

data %>% 
  ggplot(aes(unemployment, percentage20_donald_trump)) + 
  geom_point()


data %>% 
  ggplot(aes(income, percentage20_donald_trump)) + 
  geom_point()
```

# Voter Turnout Rate

```{r, load_data}

turnout_2016 <- read.csv("data/turnout_2016.csv") %>% 
  clean_names() 

turnout_2020 <- read.csv("data/turnout_2020.csv") %>% 
  clean_names()

#join two dataset
data_turnout <- turnout_2016 %>% 
  left_join(turnout_2020, by = c("state" = "state")) %>% 
  rename(turnout_2016 = turnout_rate.x, turnout_2020 = turnout_rate.y) %>% 
  select(state, turnout_2016, turnout_2020)

glimpse(data_turnout)

```

```{r, load_time_series_data}
library(extrafont)
extrafont::loadfonts(device="pdf")
turnout_1920_2020 <- read.csv("data/turnout_1920_2020.csv") %>% 
  clean_names() 

highlight <- turnout_1920_2020[turnout_1920_2020$year == 2020, ]
turnout_1920_2020 %>% 
ggplot(aes(x=year, y=united_states_presidential_vep_turnout_rate))+
  geom_line(size = 1, color = "#2c7fb8") +
  geom_point(color = "#2c7fb8") +
  geom_label(data = highlight, 
                   aes(label = "66.7%"), 
                   color = "#2c7fb8",
                   box.padding = 0.25,
                   point.padding = 0.5,
             vjust = -0.5,
             hjust = 0.6,
             family = "Courier")+
  theme_minimal() +
  labs(title = "The U.S. experienced highest turnout rate in over a century",
       subtitle = "1920-2020 Voting eligible population turnout rates",
       x = "Election Year",
       y = "Turnout Rate %") +
#  geom_vline(xintercept = 2020, color = "red", linetype = 5) +
#  gghighlight::gghighlight(year == 2020, label_key = united_states_presidential_vep_turnout_rate) +
  theme_minimal() +
  scale_x_continuous(n.break = 10) +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        text = element_text(family="Courier")) +
  expand_limits(y = c(40, 80))

```

In 2020 US election, More Americans voted than in any other in over 100 years. 66.7% percent of the voting-eligible population cast a ballot, delivering the popular vote and electoral college to Joe Biden, the Democratic candidate. Despite that the pandemic introduced a number of complications to voting day, early voting and mail-in ballots brought a record turnout for some states. 

Most of states saw an increase in voter turnout since 2016, with key battleground states like Florida, Michigan, Wisconsin and Pennsylvania seeing participation well above the national rate.

```{r}

swing_states <- c("Arizona","Texas","Florida","Georgia","Pennsylvania","Ohio","Wisconsin","North Carolina","South Carolina","Iowa","Nevada","Michigan")

swing_win <- c( "#2E74C0", "#CB454A", "#CB454A", "#2E74C0", "#2E74C0", "#CB454A", "#2E74C0","#CB454A", "#CB454A", "#CB454A",  "#2E74C0", "#2E74C0")

swing_colors <- data.frame(swing_states, swing_win)
swing_colors1 <- swing_colors$swing_win
names(swing_colors1) <- swing_colors$swing_states

data_turnout1$value <- round(data_turnout1$value,1)

ggplot(data=data_turnout1,aes(x = turnout_rate, y = value, group = state)) +
  geom_line(size = 1)+
  geom_point() +
  geom_text(aes(x = turnout_rate, y = value, label = value), vjust = 2, size = 3, family = "Courier")+
  facet_wrap(~state) +
  scale_y_continuous(limits = c(40, 80)) +
  aes(color = state) +
  scale_color_manual(values = swing_colors1) +
    theme_bw()+
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold", size = 15),
          legend.position = "none",
          axis.ticks.y = element_blank(),
          axis.text.y = element_blank(),
          strip.text.x = element_text(size = 9, face = "bold"),
          text = element_text(family="Courier"))+
  scale_x_discrete(labels=c("turnout_2016" = "2016", "turnout_2020" = "2020")) +
    labs(title = "Turnout on the rise in key battlegrounds",
         subtitle = "From 2016 to 2020 presidential election")
```



```{r, map}
data_turnout2 <- data_turnout %>% 
  mutate(change = case_when(
    turnout_2020 >= turnout_2016 ~ "increase",
    turnout_2020 < turnout_2016 ~ "decrease"))

states_sf <- get_urbn_map("states", sf = TRUE)

data_state <- states_sf %>% 
  left_join(data_turnout2, by = c("state_name" = "state"))
glimpse(data_state)

```



```{r}


data_state %>% 
  ggplot(aes()) +
  geom_sf(aes(fill = turnout_2020), colour = "#ffffff")+
  theme_void() +
  labs(title = "Key battleground states see high turnout rate",
       subtitle = "2020 Election",
       fill = "turnout rate") +
  geom_sf_text(aes(label = state_abbv), color = "white", size = 2, family = "Courier") +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 14),
        text = element_text(family="Courier")) +
  scale_fill_gradient(low = "#1a9641",
                        high = "#d7191c")
```

```{r}


    
    
data_turnout_15 <- data_turnout %>%
  arrange(desc(turnout_2020)) %>% 
  top_n(n=15, wt = turnout_2020)
glimpse(data_turnout_15)

data_turnout_15$state

top_states <- c("Minnesota","Colorado","Maine","Wisconsin","Washington","New Hampshire","Oregon","Vermont","Michigan","Iowa","Montana","Virginia","Massachusetts","Florida","North Carolina")

top_win <- c( "#2E74C0", "#2E74C0","#2E74C0", "#2E74C0", "#2E74C0", "#2E74C0", "#2E74C0", "#2E74C0", "#2E74C0", "#CB454A", "#CB454A", "#2E74C0", "#2E74C0", "#CB454A","#CB454A")

top_colors <- data.frame(top_states, top_win)
top_colors1 <- top_colors$top_win
names(top_colors1) <- top_colors$top_states



  ggplot(data = data_turnout_15, aes(x=reorder(state, turnout_2020), y = turnout_2020)) +
  geom_point(size = 2) +
  coord_flip() +
    theme_minimal() +
    expand_limits(y = c(70, 80)) +
      labs(title = "Biden won states with high turnout rate in 2020",
           subtitle = "Top 15 states with highest turnout rates",
         y = "turnout rate")+
    theme(legend.position = "none",
        axis.title.y = element_blank(),
        plot.title = element_text(face = "bold", size = 16),
        text = element_text(family="Courier")) +
  geom_text(data=data_turnout_10, aes(x=state, y=turnout_2020, label = turnout_2020), vjust = 0.2, hjust = 1.1, color = "white") +
    aes(color = state) +
    scale_color_manual(values = top_colors1)
  

```
Among swing states, Minnesota, Colorado, Maine and Wisconsin have the highest turnout rate, and Biden won the majority of votes in those states with high voter turnout rate. 

